\documentclass{article}
% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2023
\usepackage[final]{neurips}
% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{amsmath}


\title{Cooking Is All You Need: Could a Fine-Tuned SLM Outperform an LLM on the Recipe-MPR dataset?}


% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.


\author{%
  Shayan Nasiri\\
  Department of Computer Science\\
  University of Toronto\\
  \texttt{} \\
  % more authors
  \And
  Pedram Yazdinia\\
  Department of Computer Science\\
  University of Toronto\\
  \texttt{email} \\
  \AND
  Jagrit Rai\\
  Department of Computer Science\\
  University of Toronto\\
  \texttt{jagrit.rai@mail.utoronto.ca} \\
  \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}


\maketitle


\begin{abstract}
This project studies whether a small language model (SLM) can be adapted to outperform a large language model (LLM) on the Recipe-MPR dataset,
a five-way recipe recommendation question-answering task, while remaining lightweight enough for local inference. By the briefing deadline, the repository already contains a reproducible data pipeline, standardized prompt and evaluation utilities, a local general-LLM baseline through Ollama, and initial SmolLM2 prototyping for the small language model direction. The main focus so far has been to turn the project idea into a reliable
experimental foundation rather than to report final numerical results.
\end{abstract}
%
%
%
%
%
%
\section*{Attestation of Teamwork}
All three group members contributed to the project idea and framing, the comparison setup between vanilla and fine-tuned small language model and general language model, and the refinement of the repository structure.

\begin{itemize}
    \item \textbf{Pedram Yazdinia} led Phase 1 of the project by implementing the data loading and data preparation foundation. This includes validation of the raw Recipe-MPR data, normalization into a canonical processed format, deterministic train and test split generation, and the shared loading and command-line interfaces that later model experiments depend on. 

    \item \textbf{Jagrit Rai} explored and selected the initial small language model candidate, SmolLM2-135M-Instruct, and instantiated the tokenizer and model for text generation in the baseline notebook. He also documented early fine-tuning considerations for the SLM path.

    \item \textbf{Shayan Nasiri} implemented the model-facing evaluation stack for locally served language models. His contributions include the multiple choice prompt formatter, the dynamic Ollama wrapper, the response parser and evaluator, and the end-to-end local test confirming that the deepseek-r1:7b model can be queried successfully.
\end{itemize}
%
%
%
%
%
%
\newpage
\section{Problem Description}
Recipe-MPR is a multiple-choice recipe recommendation task in which each example contains a natural-language user request and five candidate recipe options, exactly one of which is correct. The dataset is challenging because the query may require more than direct keyword matching: some examples depend on common sense reasoning, some involve negation, and others include analogical or temporal cues. In the repository, these query types are tracked through the five category flags: \textit{Specific}, \textit{Commonsense}, \textit{Negated}, \textit{Analogical}, and \textit{Temporal}. 
 
The ultimate goal of the project is to determine whether a fine-tuned small language model can perform this task competitively while remaining cheaper to run and easier to inspect than a larger general-purpose model. To make this comparison meaningful, the project first needs a stable experimental foundation: fixed train/validation/test splits, a canonical processed dataset, shared prompt formatting, and a consistent way to parse and evaluate model responses.
%
%
%
%
%
%
\section{Considered Deep Learning Solution}
The team is currently pursuing two complementary modeling directions. The primary small-model direction is centered on SmolLM2-135M-Instruct, which was chosen as the initial SLM because it is compact, instruction oriented, and naturally compatible with a prompt that presents five answer options labeled A through E. This makes it a suitable candidate both for zero-shot or prompt-based baselines and for later supervised fine-tuning on the recipe task.

In parallel, the repository includes a general-LLM baseline path using models served locally through Ollama. The current code constructs a shared multiple-choice prompt, queries a selected local model, parses the returned answer letter, and computes accuracy overall and by query type. This gives the team a practical baseline that can be run locally without relying on a hosted API, and it already supports end-to-end evaluation of deepseek-r1:7b.

Beyond the notebook prototype, the repository history also shows the design progressing toward executable SLM experiments. In particular, an additional development branch contains causal-SLM prompting and configuration support around SmolLM2-style evaluation, indicating that the project has already moved from model selection into concrete experiment scaffolding.
%
%
%
%
%
%
\section{Data Acquisition}
The project uses the provided Recipe-MPR dataset stored in \textit{data/500QA.json}. This file contains 500 examples, each with a user query, five candidate recipe descriptions, a gold answer option, query-type annotations, and correctness explanations. Since every example has the same five-way structure, the task is naturally formulated as multiple-choice prediction.

Phase 1 focused on converting the raw file into a reliable shared dataset contract. The implemented preparation pipeline validates that each record contains the required fields, enforces the five-option structure, checks that the answer belongs to the option set, and preserves the original option ordering and explanation content. The pipeline then assigns stable example identifiers (rmpr-0001 through rmpr-0500) and writes a canonical processed artifact to \textit{data/processed/recipe\_mpr\_qa.jsonl}.

To support fair comparison across models, the repository also produces a deterministic split manifest at \textit{data/processed/primary\_split.json}. The split is fixed at 70\% training, 15\% validation, and 15\% test, and is stratified by the query-type signature using seed 1508. As a result, all later baselines and fine-tuned models can be evaluated on the same partition of the data rather than on ad hoc random splits.


\section{Milestones}
By the project debrief writeup, the project has advanced from an initial idea into an implemented
experimental foundation. The project repository can be found \href{https://github.com/ShayanNasiri/ece1508}{here}. 

\begin{itemize}
    \item \textbf{Data foundation completed.} The raw Recipe-MPR data has been validated, normalized, and exported into committed processed artifacts. The repository already includes deterministic split generation, reusable data loaders, and command-line utilities for preparation, validation, statistics, and split export.

    \item \textbf{Prompt and evaluation contract established.} The codebase now has a standard multiple choice prompt formatter, a parser that converts a model response back into one of the five option labels, and an evaluation script that reports exact-match accuracy overall and by query type.

    \item \textbf{Local LLM baseline operational.} The Ollama client and evaluation workflow have been implemented, and local querying of deepseek-r1:7b has been tested successfully. This means the team already has a runnable general-LLM baseline for future comparisons.

    \item \textbf{SLM direction initiated and de-risked.} SmolLM2 has been selected and instantiated for baseline text generation, and the repository history contains early causal-SLM prompting and experiment scaffolding. This reduces the uncertainty around whether the SLM direction is practically runnable.
\end{itemize}

Up until this stage, the emphasis has been on establishing a sound and reproducible design rather than on final numerical experiments. The team now has the data, prompting, and local evaluation infrastructure needed to continue fine-tuning and systematic comparison in the final report.


\end{document}